---
title: "Intro to Association Rules"
author: "KK"
date: "March 31, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#install the association rules package before you get started

#load the packages for association rules and R sample datasets

```{r}
#install.packages("arules")
#library(arules)
#library(datasets)

```


# In order to perform association rules, it is important that the data is in transactional format.
# We can also use data in the matrix format
```{r}
Groceries <- read.transactions("groceries.csv",format=c("basket"),sep=",", rm.duplicates=TRUE)

```


#plot the 'count' of each grocery item appearing in the dataset,
#from most common on down, but only show the first 20

```{r}
itemFrequencyPlot(Groceries,topN=20,type="absolute")

```

#The command to generate association rules in R is called apriori,which refers to the aprior algorithm. We will assign the algorithm's output to 'rules', which is a data frame.
#Note: you always have to specify the min support and min confidence. rules that do not have at least this support and confidence will be ignored


```{r}

rules <- apriori(Groceries,parameter=list(supp = 0.001, conf=0.8))

```

```{r}

options(digits = 2)

summary(rules)
```


# inspecting the rules
# As we observe,the output gives us the confidence, support and lift for all the rules
```{r}
inspect(rules[1:5])
```

# Now, lets identify the top rules based on "confidence"

```{r}
rules <- sort(rules,by="confidence",decreasing=TRUE)
inspect(rules[1:5])

```

#now lets generate rules where we fix some values in the LHS,say we only want to include rules where whole milk is on the LHS
#Note, %pin% allows you to do partial matching
```{r}
inspect(subset(rules, subset = lhs %pin% "whole milk" & lift >5.2))

```

# This was a basic market basket analysis where we identified which pairs of products sell well together and presence of which product increases the sales of other product
# Using the identified association rules along with taking the business context into consideration, several important business decisions can be taken.