---
title: "Outlier detection and missing value imputation"
author: "KK"
date: "March 31, 2017"
output: html_document
---

The dataset contains air pollution data from Chicago from a period Jan 1987 till end of 2005. To learn more about these measures visit http://www.airnow.gov/index.cfm?action=aqibasics.particle   

In this analysis we will be using 'dplyr' package to perform outlier analysis and we will use the 'mice' package to impute the missing values.

After imputing the missing values, using the 'dplyr' packe to perform exploratory analysis of the data

```{r}

#library(dplyr)
#library(mice)
#library(ggplot2)
#library(zoo)

```




```{r, echo=FALSE}

chicago <- readRDS("chicago.rds")

```




```{r}
head(chicago)
```

Making sure the dataset is a dplyr object and performing basic data cleansing activities that deal with outliers, data entry errors and missing values.

```{r}
chicago_d <- tbl_df(chicago)
summary(chicago_d)
```


## Outlier Analysis
```{r}
ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$tmpd)) + geom_boxplot() +labs(title="Outlier Analysis of 'temperature- tmpd'", x="chicago", y="temperature")

ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$dptp)) + geom_boxplot() +labs(title="Outlier Analysis of 'duepoint temperature - dptp'", x="chicago", y="duepoint temperature")

ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$pm10tmean2)) + geom_boxplot() +labs(title="Outlier Analysis of 'pm10tmean2'", x="chicago", y="values of pm10tmean2")

ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$pm25tmean2)) + geom_boxplot()+ labs(title="Outlier Analysis of 'pm25tmean2'", x="chicago", y="values of pm25tmean2")

ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$o3tmean2)) + geom_boxplot() +labs(title="Outlier Analysis of 'o3tmean2'", x="chicago", y="values of o3tmean2")

ggplot(chicago_d,aes(x=chicago_d$city,y=chicago_d$no2tmean2)) + geom_boxplot() +labs(title="Outlier Analysis of 'no2tmean2'", x="chicago", y="values of no2tmean2")

```

##Outlier Treatment
##replacing the outliers(values which fall outside of 3*IQR) with NA
##These NA's will be imputed using Predictive mean matching
```{r}
replace_outliers_NA <- function(x, na.rm = TRUE,...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm,... )
  H <- 3 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}


chicago_d$tmpd <- replace_outliers_NA(chicago_d$tmpd)
chicago_d$dptp <- replace_outliers_NA(chicago_d$dptp)
chicago_d$pm25tmean2 <- replace_outliers_NA(chicago_d$pm25tmean2)
chicago_d$pm10tmean2 <- replace_outliers_NA(chicago_d$pm10tmean2)
chicago_d$o3tmean2 <- replace_outliers_NA(chicago_d$o3tmean2)
chicago_d$no2tmean2 <- replace_outliers_NA(chicago_d$no2tmean2)

summary(chicago_d)
```

## Columns 'tmpd' and 'dptp' have 1 and 2 missing values only.
## Plotting the histograms to understand the distribution of the data
```{r}
g1 <- ggplot(chicago_d) + geom_histogram(aes(tmpd),binwidth = 1) 
g1 <- g1 + labs(title="Frequency distribituion of 'tmpd'", x="tmpd", y="frequency")
g1
g2 <-ggplot(chicago_d) + geom_histogram(aes(dptp),binwidth = 1)
g2 <- g2 + labs(title="Frequency distribituion of 'dptp'", x="dptp", y="frequency")
g2

```

## Both the distributuions are centered around mean with no skewness
## Using nearby values to replace these missing values

```{r}
chicago_d$tmpd <-  (na.locf(chicago_d$tmpd) + rev(na.locf(rev(chicago_d$tmpd))))/2
chicago_d$dptp <- (na.locf(chicago_d$dptp) + rev(na.locf(rev(chicago_d$dptp))))/2

```


## Columns 'pm25tmean2' and 'pm10tmean2' have high number of missing values relative to other columns
## Plotting histograms to understand the distribution of data for these two columns across all years
```{r}
g3 <- ggplot(chicago_d) + geom_histogram(aes(pm25tmean2)) 
g3 <- g1 + labs(title="Frequency distribituion of 'pm25tmean2'", x="pm25tmean", y="Count of particles")
g3
g4 <-ggplot(chicago_d) + geom_histogram(aes(pm10tmean2))
g4 <- g2 + labs(title="Frequency distribituion of 'pm10tmean2'", x="pm10tmean", y="Count of particles")
g4
```


## Using 'mice' package to replace missing values in the columns 'pm10tmean' and 'pm10tmean'
## mice uses linear regression technique to estimate the missing values using the rest of the variables as predictors 
```{r}
imputed_data <- mice(chicago_d[sapply(chicago_d,is.numeric)],m=2,maxit =10, method ='pmm')

chicago_i<- complete(imputed_data,1)
chicago_i$city <- chicago_d$city
chicago_i$date <- chicago_d$date

```

Exploratory analysis:

1. What are the days with 'seemingly' high value of PM2.5. 

Cutoff for seemingly high values has been defined based on the maximum values oserved in the past ten years
All the days with values greater than this cut off are seemigly high values
```{r}
chicago_i$year <- as.POSIXlt(chicago_i$date)$year + 1900

pm25_stats_by_year <- chicago_i %>%
  group_by(year) %>% 
  summarise(max(pm25tmean2))

previous_five <- pm25_stats_by_year %>%
   filter(year >1995)

cutoff_highpm25 <- mean(previous_five$`max(pm25tmean2)`)

seemingly_highpm25 <- chicago_i %>%
  filter(pm25tmean2 > cutoff_highpm25) %>%
  filter(year > 1995)

seemingly_highpm25 <- arrange(seemingly_highpm25,pm25tmean2)

seemingly_highpm25[1:5,]
```




Exploreing this subset of high PM2.5 level days further and see if other factors in the data could be linked to these high PM2.5 days. 

## The high PM2.5 level have a mild correlation with 'no2' and 'pm10'
```{r}
cor(seemingly_highpm25$pm25tmean2,seemingly_highpm25$pm10tmean2)
cor(seemingly_highpm25$pm25tmean2,seemingly_highpm25$o3tmean2)
cor(seemingly_highpm25$pm25tmean2,seemingly_highpm25$no2tmean2)
cor(seemingly_highpm25$pm25tmean2,seemingly_highpm25$tmpd)
cor(seemingly_highpm25$pm25tmean2,seemingly_highpm25$dptp)

ggplot(seemingly_highpm25,aes(pm10tmean2, pm25tmean2)) + geom_line(size=1)
ggplot(seemingly_highpm25,aes(o3tmean2, pm25tmean2)) + geom_line(size=1)
ggplot(seemingly_highpm25,aes(no2tmean2, pm25tmean2)) + geom_line(size=1)
ggplot(seemingly_highpm25,aes(tmpd, pm25tmean2)) + geom_line(size=1)
ggplot(seemingly_highpm25,aes(dptp, pm25tmean2)) + geom_line(size=1)

```



Key summary statistics of the pollutants by year 
```{r}

chicago_i$year <- as.POSIXlt(chicago_i$date)$year + 1900


pm25_stats_by_year <- chicago_i %>%
  group_by(year) %>% 
  summarise(mean(pm25tmean2),median(pm25tmean2),max(pm25tmean2),min(pm25tmean2))

pm10_stats_by_year <- chicago_i %>%
  group_by(year) %>% 
  summarise(mean(pm10tmean2),median(pm10tmean2),max(pm10tmean2),min(pm10tmean2))

pm10_stats_by_year <- chicago_i %>%
  group_by(year) %>% 
  summarise(mean(o3tmean2),median(o3tmean2),max(o3tmean2),min(o3tmean2))

pm10_stats_by_year <- chicago_i %>%
  group_by(year) %>% 
  summarise(mean(no2tmean2),median(no2tmean2),max(no2tmean2),min(no2tmean2))


```


Quintile analysis

Using the quintiles is a good way to understand how one variable across the quintiles of another variable

What are the average levels of Ozone (o3) within quintiles of PM25? 


```{r}
ApplyQuintiles <- function(x) {
  cut(x, breaks=c(quantile(chicago_i$pm25tmean2, probs = seq(0, 1, by = 0.20))),labels=c("Q1","Q2","Q3","Q4","Q5"), include.lowest=TRUE)
}

chicago_i$pm25.qunit <- sapply(chicago_i$pm25tmean2,ApplyQuintiles)

o3_by_pm25quantile <- chicago_i %>%
  group_by(pm25.qunit) %>%
  summarise(mean(o3tmean2))

o3_by_pm25quantile
```


Findings:

## The Ozone level across the PM25 quantiles tends to decrease with increase in PM25 until quantile 3 but this behaviour is not observed in quantile 4 where a sudden in ozone level is observed. This behaviour can  probably be better explained using other additional factors.
