---
title: "Prinicipal component analysis in R"
author: "KK"
date: "October 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
#install.packages('ggfortify')
#library(ggfortify)
#library(stats)
```

PCA constructs new axes (princpal components) in an advantageous manner: my new axes or orthogaonal--uncorrelated with each other--and a smaller number of my
new axes will capture a large amount the variance present across all my original data axes.

## PCA and Iris Data
iris dataset for a demonstration of how to use (and visualize) PCA results. 
```{r}
data("iris")
str(iris)
summary(iris)
head(iris)

```

We observe that only the data contain four continuous variables which corresponds to physical measures of
flowers and a categorical variable describing the flowers’ species.

```{r}
# pull appart continuous vs categorical values 
ir.cont <- iris[, 1:4]
ir.species <- iris[, 5]
```

Next we use the `prcomp` function, which returns an object of class prcomp, which have some methods available.
The print method returns the standard deviation of each of the four PCs, and their rotation (or loadings), which
are the coefficients of the linear combinations of the continuous variables.
```{r}
# apply PCA - scale. = TRUE is highly advisable, but default is FALSE. 
ir.pca <- prcomp(ir.cont,
                 center = TRUE,
                 scale. = TRUE)
# print method
print(ir.pca)
```

The plot method returns a plot of the variances (y-axis) associated with the PCs (x-axis).
The Figure below is useful to decide how many PCs to retain for further analysis.
In this simple case with only 4 PCs this is not a hard task and we can see that the first
two PCs explain most of the variability in the data.
```{r}
# plot method
plot(ir.pca, type = "l")
```

The summary method describe the importance of the PCs. The first row describe again
the standard deviation associated with each PC. The second row shows the proportion
of the variance in the data explained by each component while the third row describe
the cumulative proportion of explained variance. We can see there that the first two
PCs accounts for more than {95\%} of the variance of the data.
```{r}
# summary method
summary(ir.pca)
```

Next, we will project the data on the first two PCs and visualize them
```{r}
autoplot(ir.pca)
```

If you want to colorize by non-numeric values in the original data, pass original
data using `data` keyword and then specify column name by colour keyword.
Use `help(autoplot.prcomp)` (or `help(autoplot.*)` for any other objects) to check
available options.
```{r}
autoplot(ir.pca, data = iris, colour = 'Species')
```

Passing `label = TRUE` draws each data label using `rownames`
```{r}
autoplot(ir.pca, data = iris, colour = 'Species', label = TRUE, label.size = 3)
```

Passing `shape = FALSE` makes plot without points. In this case, `label` is turned on
unless otherwise specified.
```{r}
autoplot(ir.pca, data = iris, colour = 'Species', shape = FALSE, label.size = 3)
```

Passing `loadings = TRUE` draws eigenvectors.
```{r}
autoplot(ir.pca, data = iris, colour = 'Species', loadings = TRUE)
```


You can attach eigenvector labels and change some options.
```{r}
autoplot(ir.pca, data = iris, colour = 'Species',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
```

This can also all be accomplished with the `ggbiplot` function, which also gives your more
plotting flexibility.
```{r}
library(devtools)
install_github("ggbiplot", "vqv")
 
library(ggbiplot)
g <- ggbiplot(ir.pca, 
              groups = ir.species, ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)
```

The previous plot colors each point according to the flowers’ species and draws a Normal distribution
contour line with ellipse.prob probability (default to {68\%}) for each group. 

You can find more info about ggbiplot can be obtained by the usual `help(ggbiplot)`.

Also, with ggplot, other PCs can be chosen through the argument choices of the function. 
```{r}
g1 <- ggbiplot(ir.pca, 
              groups = ir.species, ellipse = TRUE, 
              circle = TRUE)
g1<- g1 + scale_color_discrete(name = '')
g1 <- g1 + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g1)
```
