---
title: "Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this regression, I used the Walmart Kaggle data set to understand relation between sales of a walmart store and different location based factors like temperature, fuel price, CPI etc
##Data Reading and Cleaning
```{r}
fullData<-read.csv('Walmart-Yearly-Sales-by-Store.csv',header = TRUE,na='NA')
summary(fullData)

# Dropping NAs to account for last year sales.  Adequate data to not negatively impact the model.
testData<-na.omit(fullData)

testData$Type <- factor(testData$Type)
testData$Store<-factor(testData$Store)
testData$Date <- as.Date(testData$Date, "%m/%d/%Y")
summary(testData)
testData$Fuel_by_cents<-testData$Fuel_Price*100
```

Rqw Data scatter
```{r}
attach(testData)

```


Scatter plots for each variable
```{r}
#Relationship between Size & Sales
plot(Size,Sales,pch=16)
abline(lm(Sales~Size),col='red',lty=2)

#Relationship between Sales & Last Year Sales
plot(LastYearSales,Sales,pch=16)
abline(lm(Sales~LastYearSales),col='red',lty=2)

#Relationship between Sales & Fuel Price
plot(Fuel_Price,Sales,pch=16)
abline(lm(Sales~Fuel_Price),col='red',lty=2)

#Relationship between Sales & Temperature
plot(Temperature,Sales,pch=16)
abline(lm(Sales~Temperature),col='red',lty=2)

#Relationship between Sales & CPI
plot(CPI,Sales,pch=16)
abline(lm(Sales~CPI),col='red',lty=2)

#Relationship between Sales & Unemployment
plot(Unemployment,Sales,pch=16)
abline(lm(Sales~Unemployment),col='red',lty=2)
```

Final model:

```{r}
finalfit<-lm(Sales ~ LastYearSales + Fuel_by_cents+ Temperature+ Unemployment + CPI  + IsHoliday)
summary(finalfit)

#Senstivity change
finalfitWithLog<-lm(log(Sales) ~ log(LastYearSales) + log(Unemployment) + log(CPI)  + IsHoliday)
summary(finalfitWithLog)
```


Assumption Testing:
**Collinearity Check:**
```{r}
cor(testData[,3:10])
```

Residual testing:
**Standardized residual plot ‐ on fitted values**
```{r}
finalfit.stres <- rstandard(finalfit)
plot(finalfit$fitted.values, finalfit.stres, pch = 16, main = "Standardized Residual Plot", xlab = "Fitted Value", ylab = "Standardized Residuals")
abline(0,0, lty=2, col="red")
```
  
**Scatterplot Matrix**
```{r}
pairs(~ finalfit.stres+ Fuel_Price +IsHoliday+ Temperature+ Unemployment +  CPI +Size, main="Simple Scatterplot Matrix")

```

```{r}
tem<-lm(Sales~Temperature)
summary(tem)

fuel<-lm(Sales~Fuel_Price)
summary(fuel)

cpi<-lm(Sales~CPI)
summary(cpi)

unemployment<-lm(Sales~Unemployment)
summary(unemployment)

size<-lm(Sales~Size)
summary(size)

lastyearsales<-lm(Sales~LastYearSales)
summary(lastyearsales)
```


Individual Scatter plots

**Standardized residual plot ‐ on Size**
```{r}
plot(Size,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="Store Size",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")

```

**Standardized residual plot ‐ on Last Year Sales**
```{r}
plot(LastYearSales,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="Last Year Sales",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")
```

**Standardized residual plot ‐ on Fuel Price**
```{r}
plot(Fuel_Price,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="Fuel Price",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")
```

**Standardized residual plot ‐ Unemployment**
```{r}
plot(Unemployment,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="Unemployment",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")

```

**Standardized residual plot ‐ Temperature**
```{r}
plot(Temperature,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="Temperature",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")

```

**Standardized residual plot ‐ on CPI**
```{r}
plot(CPI,finalfit.stres, pch=16, main="Standardized Residual Plot", xlab="CPI",ylab="Standardized Residuals")
abline(0,0,lty=2,col="blue")

```

###Residual plot Normality Checking
**Histogram with normal curve**
```{r}
h<- hist(finalfit.stres)
x<-finalfit.stres
xfit<-seq(min(x),max(x),length=50)
yfit<-dnorm(xfit,mean = mean(x),sd=sd(x))
yfit<-yfit*diff(h$mids[1:2])*length(x)
lines(xfit,yfit,col='blue')
```

**Normal probability Plot**
```{r}
qqnorm(finalfit.stres, main = "Normal Probability Plot", xlab = "Normal Scores", ylab = "Standardized Residuals")
qqline(finalfit.stres, col = "red")
```


